A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-17 tex/codatadp.tex
Most of Section 5

  * Strategies
  * Implementation
    * mention that discriminators are their own function symbols
    * what is an SMT theory solver
    * sharing is caring (and other "refinements" on calculus)
    * theory combination more complicated (and Theorem Completeness too weak)
      * need something like the string paper
    * cooperation with Nelson-Oppen / integration into SMT solver
    * "mu" stuff
    * no "internal search"
  * Enumerator?
    * for model generation

>>>>

This section describes how the calculus in Section~\ref{sec:the-calculus} is implemented within the SMT solver CVC4.

\paragraph{Optimizations}
We first discuss several optimizations that are not reflected in the presentation of the calculus, which closely follow the approach outlined by Barrett et al in~\cite{}.
We only briefly mention these optimizations, since each applies to datatypes as well as co-datatype, and most have been presented thoroughly in~\cite{}.

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$ 
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and co-datatypes.
Also, selectors are collapsed eagerly:
if $s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor 
$\const{C}_i( \vec u )$, then we immediately infer $s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Decomp} before inferring this equality.  
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a pre-processing step.
In particular, we rewrite $t \teq s$ to $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
%In DPLL(T) \rem{CDCL{T}?},
%a SAT solver
Given a theory $T = T_1 \cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL{T}?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson-Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type, 
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either 
reports that a subset $C$ of $M_i$ is unsatisfiable in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, then a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.

The rules of the calculus from Section~\ref{sec:the-calculus} are implemented as follows.
First, for each equality, we associate an \emph{explanation} for that equality.
\rem{TODO}

Step 1 of the calculus (along with the optimization for collapsing selectors) and the \rn{Singleton} rule are performed eagerly, that is for partial satisfying assignments $M$, while step 2 and the \rn{Split} rule from step 3 are performed only for complete satisfying assignments $M$, and in that order.

  * Implementation
    * what is an SMT theory solver
    * sharing is caring (and other "refinements" on calculus)
    * theory combination more complicated (and Theorem Completeness too weak)
      * need something like the string paper
    * cooperation with Nelson-Oppen / integration into SMT solver
    * "mu" stuff
    * no "internal search"
  * Enumerator?
    * for model generation
